Replacing handwritten signatures with open electronic signature software

### 2020-05-16 paper/main.tex
Remove extra new line in paper
Add progress - student scientific conference
Add changes to paper from svk to bachelor thesis

\hyphenation{elektro-nic-kých}
\hyphenation{na-ria-dením}

\title{Replacing handwritten signatures with open electronic signature software}
\author{Jakub Ďuraš}
\iffalse % Not part of the version for ŠVK
\typprace{bachelor's}
\fi
\inytypprace{Student Scientific Conference}
\rok{2020}
\odbor{Applied Informatics}
\miesto{Košice}
\veduci{RNDr. Viliam Kačala}
\pracovisko{Institute of Computer Science}
\iffalse % Not part of the version for ŠVK
\podakovanie{
  I would like to thank my supervisor Viliam Kačala for his continuous feedback and patience with my question.

  I would also like to acknowledge help received from Pavol Sokol and Laura Rózenfeldová on legal questions, and access to various resources from Gabriel Semanišin.

  Finally, I would like to thank my family and friends who contributed in various ways to my thesis.
}
\fi

\iffalse % Not part of the version for ŠVK
\pdfzadanie{zadanie.pdf}
\fi

\abstract{
With the recent changes in the legal status of electronic signatures in the world, there is an opportunity to expand the use of electronic signatures as an alternative to handwritten signatures outside the usual areas.
Electronic signatures are becoming increasingly important in the age of electronic communication when handwritten signatures are unpractical due to the need to print, personally exchange, or send signed documents.
We devote our attention to the exploration of legal preconditions relevant for the EU, technological preconditions, review of the existing software, and proposal of our own software platform.
The result is open, user-friendly software for Windows, Mac OS X, and Linux.
Using the application, it is possible, following the eIDAS Regulation, to sign and verify the signature of any document with the legal effect of an attested signature.
Our solution supports ID cards of the Slovak Republic with the eID chip without any modifications or settings, which enables easy use outside the usual use in web applications of the state.
The application is not tied only to the Slovak eID but is ready for further use with various devices in different countries and as a platform for other types of signatures.
Another part is a website that, in addition to the possibility of downloading software and brief documentation, provides general information about electronic signatures so that the project is as accessible to the public as possible.
We concentrate on simple extensibility of the program via language translations, documentation, functionality, and general information.
}

@@

\abstrakt{
So zmenami v právnom statuse elektronických podpisov vo svete sa vytvára príležitosť rozšíriť elektronické podpisovanie ako plnohodnotnú náhradu pre vlastnoručné podpisy.
Elektronické podpisy sú dôležité čoraz viac v dobe elektronickej komunikácie, kedy sú vlastnoručné podpisy nepraktické z dôvodu nutnosti tlače, osobného kontaktu alebo zasielania podpísaných dokumentov.\\
V práci sa venujeme právnym predpokladom relevantným pre EÚ, technologickým predpokladom, prehľadu v súčasnosti dostupného softvéru a návrhu vlastnej softvérovej platformy.
Výsledkom je otvorený, používateľsky prívetivý softvér pre platformy Windows, Mac OS X a Linux.
Pomocou aplikácie je možné, v súlade s nariadením eIDAS, podpísať a overiť podpis ľubovoľného dokumentu, ktorý má rovnaký právny účinok ako osvedčený podpis.
Naše riešenie podporuje občianske preukazy Slovenskej Republiky s čipom eID bez akýchkoľvek úprav alebo nastavení, čím sa umožní ich jednoduché využitie mimo obvyklého použitia vo webových aplikáciách štátu.
Aplikácia nie je viazaná iba na slovenské eID, ale je pripravená na ďalšie použitie s rôznymi zariadeniami v rôznych štátoch a ako platforma pre iné druhy podpisov.
Súčasťou je aj webová stránka, ktorá okrem možnosti stiahnutia softvéru a stručnej dokumentácie poskytuje všeobecné informácie o elektronických podpisoch, aby bol projekt čo najviac prístupný verejnosti.
Dôraz kladieme na jednoduchú rozšíriteľnosť programu vo forme jazykových prekladov, dokumentácie, funkcionality a všeobecných informácií.
}

@@

\tableofcontents
\iffalse % Not part of the version for ŠVK
\listoffigures
\listoftables
\fi

@@

Our goal is to explore an accessible way to use electronic signatures.
In the beginning, we go over the preconditions that make it a viable alternative to the handwritten signatures.
We cover the legal preconditions relevant for the European Union and Slovakia in particular.
Also, we explain technical preconditions focusing on the cryptographic concepts like asymmetric cryptography, timestamping, hashing, and standards that comply with the law within the European Union.
Secondly, we do a review of the existing software, where we briefly cover several applications that are meant for basic use.
We compare them using several criteria to find out the strengths and weaknesses of the current implementations and identify a potential gap.
Finally, we devote our attention to our own implementation.
We explain the proposed desktop software and go into details on software engineering aspects.
Specifically, modularity, architecture, usability, localization, testing, and automation.
In the end, we compare it with the existing software and identify possible future improvements.
Another result we focus on is a supporting website and how it helps with the use of the software.
All in a way that is open for extensibility by the community.

@@

\iffalse % Not part of the version for ŠVK
TODO: Think about adding section on handwritten signatures
\fi

@@

We are considering law applicable locally in the Slovak Republic (SR).
Its law is greatly influenced by the European Union (EU), being its member since 2004, and by the rest of the world.
We can assume this is at least partially applicable outside of the SR as well.

@@

\obrazok{./figures/electronic-advanced-qualified}{The relation between electronic, advanced, and qualified signatures.}{signature-comparison}

@@

\pojem{Qualified electronic signatures} (QES) are a subset of advanced electronic signatures with more specific requirements that can be found in annex 1 of the eIDAS regulation\footnote{Available at https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1.}.
In more practical terms, such advanced electronic signatures are created with a qualified device using a qualified certificate and a qualified trust service.
Qualified, in this case, means authorized for such use by the legal authorities.
These devices, certificates, and trust services are explicitly listed by the EU. In many cases indirectly by listing the organization (and its items) from a specific country.

@@

Such right can be exercised to license the software under either proprietary or open-source license.
\iffalse % Not part of the version for ŠVK
Proprietary meaning under the exclusive legal right of the author, typically also confidential and distributed as a paid product.
Open-source meaning having its source code freely available, typically also distributed for free and without any liability (the software is provided "as is").

Motivation to license the software under an open-source license can differ, and so do such licenses.
While a proprietary license is usually made specifically for that entity and its interests, open-source licenses tend to be reused between different authors.
This means that consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
\fi
In general, we can classify the open-source licenses into two categories: \pojem{permissive} and \pojem{copyleft}.

Copyleft license, in general, requires the user to publish the modified work under the compatible (free) license.
As a result, it forces them to extend the rights they have received onto others and, in turn, somehow hinder usability in software licensed under a different license.
\iffalse % Not part of the version for ŠVK
A popular example of such license is the GNU General Public License(GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).
\fi

Permissive licenses, on the other hand, do not have such a requirement in place and are more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable, and the original copyright and license are distributed with the software.
\iffalse % Not part of the version for ŠVK
Often used permissive licenses are MIT License, Apache License, or BSD License.
\fi

@@

\iffalse % Not part of the version for ŠVK
As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are the ones using someone else's software and when someone else is using our software.
\fi
Generally speaking, the more open the license is, the higher the probability it will not prevent adoption for legal reasons.
We provide examples of categorized licenses in Figure \ref{fig:license-comparison}.

\iffalse % Not part of the version for ŠVK
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

@@

Previously mentioned standards PAdES, XAdES, and CAdES define how asymmetric cryptography, hashing, or timestamping should be used to comply with the standards.
We go over a these topics and why they are able to satisfy the requirements for QES mentioned in the Section \ref{eidas}.

@@

\iffalse % Not part of the version for ŠVK
\begin{figure}[!htb]
  \centering

  \begin{tikzpicture}
    \datavisualization [scientific axes=clean, visualize as smooth line, all axes={ticks={major={at={}}}}, y axis={label={number of keys}}, x axis={label={number of users}}]
    data [format=function] {
      var x : interval [2:125];
      func y = \value x*\value x;
    };
  \end{tikzpicture}

  \centerline{\parbox{13cm}{\sl \caption{Cubic growth of the number of keys.}}}
\end{figure}
\fi

@@

\iffalse % Not part of the version for ŠVK
\obrazok{./figures/digital-signature-scheme}{Principle of digital signatures which involves signing and verifying a message (Paar 2009). TODO: I don't have rights to this image!!!}{pubkey-scheme}
\fi

@@

In Slovakia, the private key is usually handed over to the user personally, saved on the ID card. This ID card (used together with a card reader) provides an API for safe access, management, and cryptoprocessing.
Such a device is in general called a \pojem{Hardware Security Module} (HSM), and the API relevant for this thesis is \pojem{PKCS \#11} \footnote{PKCS \#11 specification available at \url{http://docs.oasis-open.org/pkcs11/pkcs11-base/v2.40/os/pkcs11-base-v2.40-os.html}.}.

@@

\iffalse % Not part of the version for ŠVK
TODO: Mention EU LOTL
\fi

@@

\iffalse % Not part of the version for ŠVK
TODO: Add scheme of document > hash - valid signature, change in document > different hash - invalid signature.
\fi

@@

\subsection{PAdES, XAdES, and CAdES}

Extensions for PDF files, XML files, and CMS signed data for signing using advanced electronic signatures, as explained in Section \ref{eidas}, are PAdES, XAdES, and CAdES, respectively.
They are defined in the respective specifications by the ETSI in \cite{pades}, \cite{xades}, and \cite{cades}.

@@

All of them support multiple signatures applied in succession.
\iffalse % Not part of the version for ŠVK
Implementation of the PAdES is rather portable since any compatible PDF file reader or editor will work with the signatures in a similar way - signature created in one should be easily verifiable in any other. XAdES and CAdES, however, are usually tied to specific implementation or standard that is built on top of them.
\fi

\begin{table}[h!]
  \begin{tabular}{ |p{2cm}||p{2.2cm}|p{2cm}|p{3cm}|p{3.9cm}| }
    \hline
    Standard & Usable with & Format & Multi-signature & Appearance \\
    \hline
    PAdES & PDF                  & embedded         & yes, sequential    & supported            \\
    XAdES & XML, any             & XML              & yes                & depends on the usage \\
    CAdES & any                  & binary           & yes                & depends on the usage \\
    \hline
  \end{tabular}

@@

The ability to keep signed documents valid for a prolonged period of time is covered in these standards under optional \pojem{Long-Term Validation} (LTV) that allows archiving of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.

\iffalse % Not part of the version for ŠVK
\section{Software engineering}

We will consider Ian Sommerville's definition of software engineeing - "an engineering discipline that is concerned with all aspects of software production from initial conception to operation and maintenance" \cite{sommerville}.
Aspects we will focus on are design, implementation, testing and releasing, and maintenance.
Small

TODO: Give definition to the software engineering, what aspects we are going over, our focus on the engineering relevant for the open source.

@@

\iffalse % Not part of the version for ŠVK
\chapter{Goals}

\section{Software}

TODO: Basis can be formed on the requirements article https://github.com/durasj/octosign/wiki/Initial-Requirements

TODO: Basic software requirements and more specific software requirements.

TODO: Enginnering parts - architecture, testing, maintainability.

TODO: Backends

\section{Information}

TODO: Website - multilingual, editable by the community
TODO: Global status - world map
\fi

@@

We focus only on simple applications made mostly for the signing of the documents.
During our review, we have used several PDF files that we have tried to sign using QES available on the SR eID.
We have tried to include software of all types in our review, preferring the most popular, locally available ones, since those should be prepared for the use with the SR eID.
Excluded are all complex applications with advanced features and high price tag.
We have picked one open-source desktop application, one web application - Software as a Service (SaaS), one PDF viewer, and one simple paid desktop application. Criteria we focus on are:

@@

\begin{table}[h!]
  \begin{tabular}{ p{2.75cm}|p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm} }
                      & \rot{licensing} & \rot{cross-platform} & \rot{localization} & \rot{file suppport} & \rot{easy setup} & \rot{flexibility} & \rot{built-in verification} & \rot{extensibility} \\
    \hline
    JSignPdf          & \doublecheck    & \checkmark           &                    &                     &                  & \doublecheck      &                             & \checkmark \\
    zep.disig.sk      & \checkmark      & \doublecheck         & \doublecheck       & \checkmark          & \doublecheck     &                   & \doublecheck                &            \\
    Acrobat DC        & \checkmark      & \checkmark           & \doublecheck       &                     & \checkmark       & \doublecheck      & \doublecheck                &            \\
    D.PDF Signer      &                 &                      & \checkmark         &                     & \doublecheck     & \checkmark        &                             &            \\
  \end{tabular}

@@

\section{Commercial desktop application D.PDF Signer}

@@

Just as the zep.digis.sk, D.PDF signer utilizes the currently bundled software with the SR eID.
It is straightforward and flexible.
Downsides are lack of support for platforms other than Windows, localization, and built-in signature verification.

\iffalse % Not part of the version for ŠVK
\section{Modularity}

TODO: Rough overview of different ways modular architecture can be achieved. Focus on modularity via multiple independant applications.

\subsection{Communication}

TODO: Communication in multiple independant applications - IPC, Sockets, Pipes, STDIO.

\subsection{Bundling}

\subsection{Communicating and bundling with GPL software}
\fi

@@

We have developed an open-source, cross-platform desktop software with focus on the modularity, architecture, usability, localization, automated testing with overall automation, and extensibility by the community.
The source code is available at \url{https://github.com/durasj/octosign}, \url{https://github.com/durasj/octosign-dss}, and \url{https://github.com/durasj/octosign-image}.

\subsection{Usability and localization}

@@

We have also made the application easily translatable and try to display data like dates in the local format.
When the application is started, we try to detect the locale of the user and set it as a default language.
If we do not yet support the locale of the user, we default to the English language.
At the time of writing, our application is fully translated in three languages: English (source language), Slovak, and Czech.

The process of content translation is mostly automated.
We start with the automated extraction of the source strings from the code of the main module and files provided by the backends.
Next, we manually include the extracted source strings in the commit to our versioning control - we have retained manual control over this to prevent a noise that automated commits could make.
On push to the hosted repository, a localization platform Transifex\footnote{Transifex is a SaaS localization platform free for OSS, see \url{https://www.transifex.com/}.} automatically pulls and process the updates.
These phrases can then be translated by anybody using the UI of the Transifex platform to make it easily accessible.
Once the changes are made on the Transifex platform, it automatically creates a pull request on our hosted repository with the changes made to that particular language.
As we explain in detail in Section \ref{testing}, part of the build process is to incorporate these in the application.

@@

Signing modules - backends - are separate CLI applications written in any programming language.
They are responsible for the actual document manipulation - signing and verification.

The main module loads all available backends at runtime and verifies their availability, options, and supported file types.
Once the user decides to verify or sign a file, they are executed as CLI applications.

Communication between the main module and backends is in plaintext over the standard streams - standard input, output, and error.
To make sure we have correctly documented and easily accessible way for backend development, we have prepared a Backend specification\footnote{Backend specification is available at \url{https://github.com/durasj/octosign/wiki/Backend-specification}.} that specifies what is required of the backend, what is optionally supported, and the format of the inputs, outputs, and error handling.

\iffalse % Not part of the version for ŠVK
TODO: Go a bit more over the backend specification here
\fi

@@

>>>>

\hyphenation{elektro-nic-kých}
\hyphenation{na-ria-dením}
\hyphenation{možno-sťami}
\hyphenation{používateľ-sky}
\hyphenation{akých-koľ-vek}

\title{Replacing handwritten signatures with open electronic signature software}
\author{Jakub Ďuraš}
\typprace{bachelor's}
\rok{2020}
\odbor{Applied Informatics}
\miesto{Košice}
\veduci{RNDr. Viliam Kačala}
\pracovisko{Institute of Computer Science}
\podakovanie{
I want to thank my supervisor RNDr. Viliam Kačala for his continuous feedback and patience with my questions and late submissions.
I would also like to acknowledge help received from RNDr. JUDr. Pavol Sokol, PhD. on legal questions, and access to resources from prof. RNDr. Gabriel Semanišin, PhD.
Finally, I would like to thank my family and friends who contributed in various ways to my thesis.
}

\pdfzadanie{assignment.pdf}

\abstract{
With the recent changes in the legal status of electronic signatures in the world, there is an opportunity to expand the use of electronic signatures as an alternative to handwritten signatures outside the usual areas.
We devote our attention to the exploration of legal preconditions relevant for the EU, technological preconditions, review of the existing software, and proposal of our own software platform.
The result is an open software platform for operating systems Windows, Mac OS X, and Linux.
Using the application, it is possible, following the eIDAS Regulation, to sign and verify the signature of any document with the legal effect of an attested signature.
Our solution supports ID cards of the Slovak Republic with the eID chip without any modifications or settings, which enables use outside the web applications of the state.
The application is not tied only to the Slovak eID but is ready for further use with various devices in different countries and as a platform for other types of signatures.
Another part is a website that, in addition to the possibility of downloading software and brief documentation, provides general information about electronic signatures.
The emphasis is on simple extensibility of the program via language translations, documentation, functionality, and general information.
The thesis can be further extended with novel ways of signing or by improving flexibility, which is not up to par with compared software.
}

@@

\abstrakt{
So zmenami v právnom statuse elektronických podpisov vo svete sa vytvára príležitosť rozšíriť elektronické podpisovanie ako plnohodnotnú náhradu pre vlastnoručné podpisy.
V práci sa venujeme právnym predpokladom relevantným pre EÚ, technologickým predpokladom, prehľadu v súčasnosti dostupného softvéru a návrhu vlastnej softvérovej platformy.
Výsledkom je otvorená softvérová platforma pre operačné systémy Windows, Mac OS X a Linux.
Pomocou aplikácie je možné, v súlade s nariadením eIDAS, podpísať a overiť podpis ľubovoľného dokumentu, ktorý má rovnaký právny účinok ako osvedčený podpis.
Naše riešenie podporuje občianske preukazy Slovenskej Republiky s čipom eID bez akýchkoľvek úprav alebo nastavení, čím sa umožňuje ich využitie mimo použitia vo webových aplikáciách štátu.
Aplikácia nie je viazaná iba na slovenské eID, ale je pripravená na ďalšie použitie s rôznymi zariadeniami v rôznych štátoch a ako platforma pre iné druhy podpisov.
Súčasťou je aj webová stránka, ktorá okrem možnosti stiahnutia softvéru a stručnej dokumentácie poskytuje všeobecné informácie o elektronických podpisoch.
Dôraz je kladený na jednoduchú rozšíriteľnosť programu vo forme jazykových prekladov, dokumentácie, funkcionality a všeobecných informácií.
Prácu je možné ďalej rožšíriť nepreskúmanými možnosťami podpisovania či zlepšením flexibility, ktorá nie je na úrovni niektorých zástupcov porovnávaného softvéru. 
}

@@

\tableofcontents

@@

Our goal is to explore an accessible way to use electronic signatures.
In the beginning, we go over the preconditions that make it a viable alternative to the handwritten signatures.
We cover the legal preconditions relevant for the European Union and Slovakia in particular.
Also, we explain technical preconditions focusing on the cryptographic concepts like asymmetric cryptography, timestamping, hashing, and standards that comply with the law within the European Union.
Secondly, we do a review of the existing software, where we briefly cover several applications that are meant for basic use.
We compare them using several criteria to find out the strengths and weaknesses of the current implementations and identify a potential gap.
Finally, we devote our attention to our own implementation.
We explain the proposed desktop software and go into details on software engineering aspects.
Specifically, we cover usability, localization, modularity, architecture, testing, and automation.
In the end, we compare it with the existing software and identify possible future improvements.
Another result we focus on is a supporting website and how it helps with the use of the software.
All in a way that is open for extensibility by the community.

@@

@@

Unless stated otherwise, we are considering law applicable locally in the Slovak Republic (SR).
Its law is greatly influenced by the European Union (EU), being its member since 2004, and by the rest of the world.
We can assume this is at least partially applicable outside of the SR as well.
The \cite{law} shapes our overall view on the global legal status of signatures and chapters 1, 2, 3, 6, and 7 of \cite{osslicensing} our Chapter \ref{copyright}.

@@

\obrazok{./figures/electronic-advanced-qualified}{Relation between electronic, advanced, and qualified signatures.}{signature-comparison}

@@

\pojem{Qualified electronic signatures} (QES) are a subset of advanced electronic signatures with more specific requirements that can be found in annex 1 of the eIDAS regulation\footnote{Available at \url{https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1}.}.
In more practical terms, such advanced electronic signatures are created with a qualified device using a qualified certificate and a qualified trust service.
Qualified, in this case, means authorized for such use by the legal authorities.
These devices, certificates, and trust services are explicitly listed by the EU. In many cases indirectly by listing the organization (and its items) from a specific country.

@@

Such right can be exercised to license the software under either proprietary or open-source license.
Proprietary meaning under the exclusive legal right of the author, typically also confidential and distributed as a paid product.
Open-source meaning having its source code freely available, typically also distributed for free and without any liability (the software is provided "as is").

Motivation to license the software under an open-source license can differ, and so do such licenses.
While a proprietary license is usually made specifically for that entity and its interests, open-source licenses tend to be reused between different authors.
This means that consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
In general, we can classify the open-source licenses into two categories: \pojem{permissive} and \pojem{copyleft}.

Copyleft license, in general, requires the user to publish the modified work under the compatible (free) license.
As a result, it forces them to extend the rights they have received onto others and, in turn, somehow hinder usability in software licensed under a different license.
A popular example of such license is the GNU General Public License (GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).

Permissive licenses, on the other hand, do not have such a requirement in place and are more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable, and the original copyright and license are distributed with the software.
Often used permissive licenses are MIT License, Apache License, or BSD License.

@@

As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are the ones using someone else's software and when someone else is using our software.
Generally speaking, the more open the license is, the higher the probability it will not prevent adoption for legal reasons.
We provide examples of categorized licenses in Figure \ref{fig:license-comparison}.

\iffalse % I have no space for this :(
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

@@

Previously mentioned standards PAdES, XAdES, and CAdES define how asymmetric cryptography, hashing, or timestamping should be used to comply with the standards.
We go over these topics and why they are able to satisfy the requirements for QES mentioned in the Section \ref{eidas}.
We draw our knowledge in chapters about cryptographic primitives from the \cite{cryptotxtbook} and \cite{cryptojoy}.

@@

@@

@@

In Slovakia, the private key is usually handed over to the user personally, saved on the ID card. This ID card (used together with a card reader) provides an \pojem{Application Programming Inteface} (API) for safe access, management, and cryptoprocessing.
Such a device is in general called a \pojem{Hardware Security Module} (HSM), and the API relevant for this thesis is \pojem{PKCS \#11} \footnote{PKCS \#11 specification available at \url{http://docs.oasis-open.org/pkcs11/pkcs11-base/v2.40/os/pkcs11-base-v2.40-os.html}.}.

@@

Specific for the EU is online \pojem{List of The Lists} (LOTL) that contains a list with trusted service providers for each country.
Each member state is obligated to maintain such list, as can be seen in Article 22 of \cite{eidas}.
From the technical point of view, it is an XML document that is signed and can be processed automatically.

@@

@@

\subsection{PAdES, XAdES, and CAdES} \label{etsi}

Extensions for Portable Document Format (PDF) files, eXtensible Markup Language (XML) files, and Cryptographic Message Syntax (CMS) signed data for signing using advanced electronic signatures, as explained in Section \ref{eidas}, are PAdES, XAdES, and CAdES, respectively.
They are defined in the respective specifications by the ETSI in \cite{pades}, \cite{xades}, and \cite{cades}.

@@

All of them support multiple signatures applied in succession.
Implementation of the PAdES is rather portable since any compatible PDF file reader or editor will work with the signatures in a similar way - signature created in one should be easily verifiable in any other. XAdES and CAdES, however, are usually tied to specific implementation or standard that is built on top of them.

\begin{table}[!ht]
  \begin{tabular}{ |p{2cm}||p{2.2cm}|p{2cm}|p{3cm}|p{3.9cm}| }
    \hline
    Standard & Usable with & Format & Multi-signature & Appearance \\
    \hline
    PAdES & PDF                  & embedded         & yes, sequential    & supported            \\
    XAdES & XML, any             & XML              & yes                & depends on the usage \\
    CAdES & any                  & binary           & yes                & depends on the usage \\
    \hline
  \end{tabular}

@@

The ability to keep signed documents valid for a prolonged period of time is covered in these standards under optional \pojem{Long-Term Validation} (LTV) that allows archiving of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.
Different profiles - levels - for different standards can be seen in Table \ref{table:2}.

\begin{table}[!ht]
  \centering
  \begin{tabular}{|l||l|l|l|}
    \hline
    Level                     & XAdES     & CAdES     & PAdES     \\
    \hline
    Basic                     & XAdES-B   & CAdES-B   & AdES-B    \\
    B with Timestamp          & XAdES-T   & CAdES-T   & PAdES-T   \\
    T with Long Term Data     & XAdES-LT  & CAdES-LT  & PAdES-LT  \\
    LT with Archive timestamp & XAdES-LTA & CAdES-LTA & PAdES-LTA \\
    \hline
  \end{tabular}

  \caption{Different profiles - levels - for standards PAdES, XAdES, and CAdES.}
  \label{table:2}
\end{table}

\iffalse % I have no space for this :(
\section{Software engineering}

We will consider Ian Sommerville's definition of software engineeing - "an engineering discipline that is concerned with all aspects of software production from initial conception to operation and maintenance" \cite{sommerville}.
Aspects we will focus on are design, implementation, testing and releasing, and maintenance.
We try to consider specifics of the open-source software as they are sometimes defined in Chapter 7 of \cite{oss}.

@@

@@

We focus only on simple applications made mostly for the signing of the documents.
During our review, we have used several PDF files that we have tried to sign using QES available on the SR eID.
We have tried to include software of all types in our review, preferring the most popular, locally available ones, since those should be prepared for the use with the SR eID.
Excluded are all complex applications with advanced features and high price tag.
We have picked one open-source desktop application, one web application - \pojem{Software as a Service} (SaaS), one PDF viewer, one simple paid desktop application, and one commercial desktop application with free version. Criteria we focus on are:

@@

\begin{table}[h!]
  \begin{tabular}{ p{2.75cm}|p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm} }
                      & \rot{licensing} & \rot{cross-platform} & \rot{localization} & \rot{file suppport} & \rot{easy setup} & \rot{flexibility} & \rot{built-in verification} & \rot{extensibility} \\
    \hline
    JSignPdf          & \doublecheck    & \checkmark           &                    &                     &                  & \doublecheck      &                             & \checkmark \\
    zep.disig.sk      & \checkmark      & \doublecheck         & \doublecheck       & \checkmark          & \doublecheck     &                   & \doublecheck                &            \\
    Acrobat DC        & \checkmark      & \checkmark           & \doublecheck       &                     & \checkmark       & \doublecheck      & \doublecheck                &            \\
    D.PDF Signer      &                 &                      & \checkmark         &                     & \doublecheck     & \checkmark        &                             &            \\
    Podpisuj.sk       & \checkmark      & \doublecheck         & \checkmark         & \doublecheck        & \doublecheck     & \checkmark        & \doublecheck                &            \\
  \end{tabular}

@@

\section{Commercial application D.PDF Signer}

@@

Just like the zep.digis.sk, D.PDF signer utilizes the currently bundled software with the SR eID.
It is straightforward and flexible.
On the other thand, it lacks support for platforms other than Windows.
Additionally, the application is only partially localized to English.
It does not support file formats other than PDF and has no built-in signature verification.

\section{Commercial desktop application Podpisuj.sk}

Application Podpisuj.sk is a paid proprietary desktop application with a free version developed locally in Slovakia.

\obrazok{./figures/screenshot-podpisujsk}{Screenshot of the Podpisuj.sk.}{screenshot-podpisujsk}

The Podpisuj.sk can be downloaded for all three major desktop platforms.
It does not rely on the software bundled with the SR eID.
It is relatively easy to use and comes fully preconfigured so it can be used with the SR eID right away.
Flexibility is not as great as with some other applications we have reviewed, and some options (like custom TSP) require one of the paid versions.
Not everything is fully localized - especially the website, and any documentation we could find was available only in Slovak.
It is not possible to extend it, but it is easy to use different tokens for signing.

@@

We have developed an open-source, cross-platform desktop software with focus on the modularity, architecture, usability, localization, automated testing with overall automation, and extensibility by the community.
Our knowledge on these topics comes from Ian Sommerville's \cite{sommerville}.
We try to consider specifics of the open-source software as they are sometimes defined in Chapter 7 of \cite{oss}.

\subsection{Usability}

@@

\subsection{Localization}

We have also made the application easily translatable and try to display data like dates in the local format.
When the application is started, we try to detect the locale of the user and set it as a default language.
If we do not yet support the locale of the user, we default to the English language.
At the time of writing, our application is fully translated in three languages: English (source language), Slovak, and Czech.
Additional languages - German and Spanish - are actively being developed.

The process of content translation is mostly automated.
We start with the automated extraction of the source strings from the code of the main module and files provided by the backends.
Next, we manually include the extracted source strings in the commit to our versioning control - we have retained manual control over this to prevent a noise that automated commits could make.
On push to the hosted repository, a localization platform Transifex\footnote{Transifex is a SaaS localization platform free for OSS, see \url{https://www.transifex.com/}.} automatically pulls and process the updates.
These phrases can then be translated by anybody using the \textit{User Interface} (UI) of the Transifex platform to make it easily accessible.
Once the changes are made on the Transifex platform, it automatically creates a pull request on our hosted repository with the changes made to that particular language.
As we explain in detail in Section \ref{testing}, part of the build process is to incorporate these in the application.

@@

Signing modules - backends - are separate \pojem{Command Line Interface} (CLI) applications written in any programming language.
They are responsible for the actual document manipulation - signing and verification.

The main module loads all available backends at runtime.
Once the user decides to verify or sign a file, they are executed as CLI applications.

Communication between the main module and backends is in plaintext over the standard streams (STDIO) - standard input (STDIN), output (STDOUT), and error (STDERR).
To make sure we have correctly documented and easily accessible way for backend development, we have prepared a Backend specification that specifies what is required of the backend, what is optionally supported, and the format of the inputs, outputs, and error handling.

@@

\subsection{Signing backends} \label{backends}

Following our specification, each backend is required to provide a configuration file and follow the communication protocol.

Configuration files are written in YAML\footnote{YAML is a human-readable data serialization language, see \url{https://yaml.org/}.} and contain basic information about the backend like name, description, repository, version, author, or license.
It also contains two properties defining what command should be executed and with what parameters to execute the backend, and build the backend, respectively.

Backend is always called with operation as the first argument, followed by an additional argument that differs between operations.
The backend should be stateless - that means it should not be influenced by any state it would save like info about what files it signed.
All communication is in UTF-8 encoded plaintext.
It needs to implement operations \textit{meta}, \textit{sign}, and \textit{verify} that are run using the configured executable, for example, \textit{./backend-executable meta}, or, \textit{./backend-executable sign /home/path.pdf}.

Operation meta is called to get info about its availability, options, and supported file types.
This allows the backend to check the platform, available software of the user, etc. and make the decision based on that.

Operation sign and verify are called to sign or verify the file with the absolute path to the file.
The main module never sends anything over the STDIO unless it's asked to do so from the backend.
That means the backend acts as a \textit{master} and the main module as a \textit{slave}.
The UI can end the process if any unexpected situation arises and displays an error.
The backend can during those two operations, at any moment: write error to the STDERR that is immediately displayed; prompt for additional information by writing to the STDOUT; get the current value of any option by writing to the STDOUT.

Ending the process with the exit code 0 signals success.
Some operations expect a result on the STDOUT before the process exits.
If code other than 0 is used, the backend should always send an error message to the STDERR.

Messages exchanged via the STDIN and STDOUT can be sometimes polluted by different libraries used on either part, so they are distinguished from the rest by using the following delimiter:

\begin{verbatim}
---TYPE---
any information exchanged in the appropriate format
---TYPE---
\end{verbatim}

\textit{TYPE} here is replaced by \textit{PROMPT}, \textit{GETOPTION}, or \textit{RESULT} based on the type of the message.
Such messages should be written in one chunk and flushed. Communication on the STDERR is accepted and displayed as-is.

Prompt is used to interactively ask for more information from the user.
The backend can ask to get an absolute path to the output file, absolute path to opened file, text, password, image (drawn or picked), or position and width of the signature that will be placed in the UI by the user.

Getting an option is used to ask for the current option value that can be set in the UI on the Settings screen.
The option is automatically prefixed on the main module for each backend so it can not retrieve options from other backends to prevent collisions and security issues.

Such communication protocol allows using any language that can simply work with the STDIO without a need for knowledge or third-party libraries to work with serialized data in formats like JSON.
The backend does not have to use any regular expressions but always simply reads the whole line.
It is supposed to be as simple as possible for newcomers.

The current version of the full specification is available at \url{https://github.com/durasj/octosign/wiki/Backend-specification}.

\subsubsection{DSS eIDAS backend}

The backend that we call \pojem{Advanced Electronic Signature} (AES), is a Java CLI application that uses \pojem{Digital Signature Service} (DSS) framework. 
The DSS is a project from the Connecting Europe Facility (CEF) Digital, a European Union fund for pan-European infrastructure investment.
It is one of the \textit{}{Building Blocks}, as they are called on their website, which helps to create and verify electronic signatures in line with European standards.
Its documentation, available in \cite{ecdss}, also provides general information about the used standards.

Source of the private key is configurable, and we can use PKCS \#11 to utilize those in HSMs, PKCS \#12 to work with key stores in files, or Microsoft Crypto API that provides an abstraction on different sources.
It supports all standards mentioned in Chapter \ref{etsi} in various configurations.
Since we are trying to simplify it down for the user, when signing, we automatically choose to use PAdES when signing PDF, XAdES when signing XML with enveloped signature, and ASiC with CAdES for all other file formats.
This could be further extended in the future, so the user can manually choose the preferred standard, container type, or signature policy.

Signature verification, however, works with all different combinations that are detected automatically.
During the verification, we construct a trusted list of root certificates from the LOTL.
Different lists for different countries are hosted individually on each countries servers.
For better performance and reliability when there is an outage on those servers, we proxy requests through CloudFlare\footnote{CloudFlare is a CDN provider, see \url{https://www.cloudflare.com/}.} \pojem{Content Delivery Network} (CDN) to distribute cached files that are periodically downloaded and stored on the Amazon Web Services\footnote{Amazon Web Services is a cloud provider with various services, see \url{https://aws.amazon.com/what-is-aws/}.} (AWS) infrastructure using the AWS Lambda and AWS S3.
